---
title: "ASSIGNMENT 06.2"
author: "Shubha, Adithya"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
subtitle: Query Tabular Data with SQL
---

```{r,echo=FALSE,warning=FALSE}
# Install program required packages =============================
# Installs sqldf,RSQLite and kableExtra
# Referred from: http://artificium.us/lessons/06.r/l-6-104-r4progs/l-6-104.html#Install_Packages_on_Demand
installRequiredPackages <- function() {
  packages <- c("sqldf","RSQLite","kableExtra")
  # Install packages that are not installed
  installed_packages <- packages %in% rownames(installed.packages())
  if (any(installed_packages == FALSE)) {
    install.packages(packages[!installed_packages])
  }
  
  suppressMessages({
  library(sqldf)
  library(RSQLite)
  library(kableExtra)
})
}  
installRequiredPackages()  
```

## Analysis of Restaurant Visits
```{r ,echo=FALSE,  warning=FALSE}

# reading the csv file from s3 bucket.
dfJanToMarch <- read.csv(file = "https://s3.us-east-2.amazonaws.com/artificium.us/assignments/80.xml/a-80-305/gen-xml/synthsalestxns-Jan2Mar.csv", header = TRUE, stringsAsFactors = FALSE)

dfSeptToOct <- read.csv(file = "https://s3.us-east-2.amazonaws.com/artificium.us/assignments/80.xml/a-80-305/gen-xml/synthsalestxns-Sep2Oct.csv", header = TRUE, stringsAsFactors = FALSE) 

dfNovToDec <- read.csv(file = "https://s3.us-east-2.amazonaws.com/artificium.us/assignments/80.xml/a-80-305/gen-xml/synthsalestxns-Nov2Dec.csv", header = TRUE, stringsAsFactors = FALSE) 

# combining all three csv files to perform analysis.
# referredFrom: https://www.w3schools.com/sql/sql_union.asp
combineDFQuery <- "SELECT * FROM dfJanToMarch
UNION ALL
SELECT * FROM dfSeptToOct
UNION ALL
SELECT * FROM dfNovToDec;
"

# quering from sqldf was referred from: http://artificium.us/lessons/06.r/l-6-301-sqlite-from-r/l-6-301.html#Querying_Data_Frames_with_sqldf
dfCombined <- sqldf(combineDFQuery)

# removing dollar symbol from amount and converting it to numeric. 
# referredFrom: http://artificium.us/lessons/06.r/l-6-112-text-proc/l-6-112.html
dfCombined$amount <- as.numeric(gsub("\\$", "", dfCombined$amount))

# total visits will be the number of rows present after combining the data
# Each row with name and guest is assumed to be 1 visit.
totalVisitQuery <- "SELECT COUNT(*) AS totalVisit FROM dfCombined;"
totalVisit <- sqldf(totalVisitQuery)$totalVisit

distinctRestaurantQuery <- "SELECT COUNT(DISTINCT restaurant) AS distinctRestaurant FROM dfCombined;"

distinctRestaurants <- sqldf(distinctRestaurantQuery)$distinctRestaurant

# using SQL inbuild function to compute median
medianQuery <- "SELECT ROUND(MEDIAN(amount),2) AS amountMedian FROM dfCombined;"
amountMedian <- sqldf(medianQuery)$amountMedian

# Computing mean to compute standard deviation

meanQuery <- "SELECT ROUND(AVG(amount),2) AS meanAmount FROM dfCombined;"
meanAmount <- sqldf(meanQuery)$meanAmount

# Formula of standard deviation referredFrom: https://www.cuemath.com/data/standard-deviation/
varianceQuery <- sprintf("SELECT AVG((amount - %s) * (amount -%s)) AS variance FROM dfCombined;", meanAmount, meanAmount)
variance <- sqldf(varianceQuery)$variance
amountStdDev <- round(sqrt(variance), 2)


restaurantWithLeastVisitQuery <- "SELECT restaurant
    FROM dfCombined
    GROUP BY restaurant
    ORDER BY  COUNT(*) ASC
    LIMIT 1;"
restaurantWithLeastVisit <- sqldf(restaurantWithLeastVisitQuery)$restaurant

# case in sql referred from:  https://www.w3schools.com/sql/sql_case.asp
groupsConsumingWineQuery <- "SELECT 
    SUM(CASE WHEN wine = 'Yes' THEN 1 ELSE 0 END)  AS consumedWine
  FROM dfCombined;"

groupsConsumingWine <- sqldf(groupsConsumingWineQuery)$consumedWine

percentageOfGroupsConsumingWine <- round((groupsConsumingWine*100)/totalVisit,2)

# Adding guest+1 assuming that guest column denotes the guest bought along
# and name column is the actual customer
averagePartySizeQuery <- "SELECT ROUND(AVG(guests + 1),2) as averagePartySize FROM dfCombined;"
averagePartySize <- sqldf(averagePartySizeQuery)$averagePartySize
```
We recorded a total number of `r totalVisit` visits across `r distinctRestaurants` different restaurants.The median amount spent for a visit across all restaurants is `r amountMedian` (σ = `r amountStdDev`).The least number of visits visits were to `r restaurantWithLeastVisit`.The average amount spent in a visit was `r meanAmount` and `r percentageOfGroupsConsumingWine`% of all parties consumed wine.The average size of a party was `r averagePartySize`.


## Summary of Revenue
```{r ,echo=FALSE}
dfCombined$date <-  as.Date(dfCombined$date, format = "%m/%d/%y")
# creating a new column with only month.
dfCombined$month <- format(dfCombined$date, '%m')

summaryOfRevenueQuery <-  "
    SELECT 
           month,
           ROUND(SUM(amount),2) AS totalRevenue,
           COUNT(*) AS totalVisits
    FROM dfCombined 
    GROUP BY month;"

summaryOfRevenue <- sqldf(summaryOfRevenueQuery)
# Creating a dataframe from 1,12 to check if all months are present or not.
# referredFrom: https://www.w3schools.com/r/r_data_frames.asp
allMonths <-data.frame(month = 1:12)

# formatting all months to readable format. 
# left join ensures that month not present in combinedDf will result in null.
# using COALESCE to remove null and replace it with -- after joining. 
# referredFrom: https://www.w3schools.com/sql/func_sqlserver_coalesce.asp
summaryForAllMonthsQuery <-  
"Select CASE 
    WHEN M.month = 1 THEN 'January'
    WHEN M.month = 2 THEN 'February'
    WHEN M.month = 3 THEN 'March'
    WHEN M.month = 4 THEN 'April'
    WHEN M.month = 5 THEN 'May'
    WHEN M.month = 6 THEN 'June'
    WHEN M.month = 7 THEN 'July'
    WHEN M.month = 8 THEN 'August'
    WHEN M.month = 9 THEN 'September'
    WHEN M.month = 10 THEN 'October'
    WHEN M.month = 11 THEN 'November'
    WHEN M.month = 12 THEN 'December'
  END AS month,
COALESCE(SR.totalRevenue, '--') as totalRevenue,
COALESCE(SR.totalVisits, '--') as totalVisits 
FROM allMonths M LEFT JOIN summaryOfRevenue SR ON
M.month = SR.month
ORDER BY M.month
"
summaryForAllMonths <- sqldf(summaryForAllMonthsQuery)
# referredFrom: https://www.rdocumentation.org/packages/knitr/versions/1.48/topics/kable
kable(summaryForAllMonths, 
      col.names = c("Month", "Total Revenue", "Total Visits"),
      format = "html") %>%
# referredFrom: https://www.rdocumentation.org/packages/kableExtra/versions/1.4.0/topics/kable_styling
kable_styling(full_width = FALSE)
```

