---
title: "Sales Analysis"
author: "Shubha, Adithya"
email: "shubha.a@northeastern.edu"
date: "2024-12-06"
output:
  html_document:
    df_print: paged
subtitle: 2020-2023
---

```{r cleanEnvironment, echo=FALSE,warning=FALSE, eval= TRUE}
# Cleans the environment before beginning the execution
# referredFrom:https://northeastern.instructure.com/courses/192346/assignments/2351524
rm(list = ls())
```

```{r installAndLoadRequiredPackages, echo=FALSE,warning=FALSE, eval= TRUE}
# Install program required packages
# Installs RMysql, DBI and testthat,kableExtra,jsonlite packages
# referredFrom: http://artificium.us/lessons/06.r/l-6-104-r4progs/l-6-104.html#Install_Packages_on_Demand
installRequiredPackages <- function() {
  packages <- c("RMySQL", "DBI","kableExtra")
  # Install packages that are not installed
  installed_packages <- packages %in% rownames(installed.packages())
  if (any(installed_packages == FALSE)) {
    install.packages(packages[!installed_packages])
  }
}

# Loads the required packages to the environment
# Loads RMysql, DBI,kableExtra and testthat packages
loadRequiredPackages <- function() {
  suppressMessages({
    library(RMySQL)
    library(DBI)
    library(kableExtra)
  })
}
installRequiredPackages()
loadRequiredPackages()
```


```{r connectToDatabase, echo=FALSE,warning=FALSE, eval= TRUE}
# Part 3: Connecting to the database

# Function to connect to the database
# @return: database connection
# referredFrom: http://artificium.us/lessons/06.r/l-6-301-sqlite-from-r/l-6-301.html#Connect_to_Database
connectToDatabase <- function() {
  # DB credentials
  dbName <- "defaultdb"
  dbUser <- "avnadmin"
  dbPassword <- "AVNS_LfgR_-4_x2xF-vv7_Bg"
  dbHost <- "practicum-1-cs-5200.b.aivencloud.com"
  dbPort <- 22694
  
  
  return (
    dbConnect(
      RMySQL::MySQL(),
      user = dbUser,
      password = dbPassword,
      dbname = dbName,
      host = dbHost,
      port = dbPort
    )
  )
}
```

```{r globalVariables, echo=FALSE,warning=FALSE, eval=TRUE}
# Define all the global variables used in the code

# Global variable to connect to the database
dbCon <- connectToDatabase()
```


## Sales Analysis
```{r getSalesData, echo=FALSE,warning=FALSE, eval=TRUE}
# Function to get the year information from database
# @param dbCon: database connection
getYearInformation <- function(dbCon) {
  # Query to get the minimum and maximum year from the sales table
  query <- "SELECT  DISTINCT product_facts.year
FROM product_facts
ORDER BY year DESC"
  
  # Execute the query
  return (dbGetQuery(dbCon, query))
}

yearInformation <- getYearInformation(dbCon)
maxYear <- yearInformation$year[1]
minYear <- yearInformation$year[length(yearInformation$year)]
```
The top five products with the most sales/revenue across all years from `r minYear` to 
`r maxYear` are shown in the table below.

```{r topFiveProducts, echo=FALSE,warning=FALSE, eval=TRUE}
displayReport <- function(dataFrame,columns){
  # referredFrom: https://www.rdocumentation.org/packages/knitr/versions/1.48/topics/kable
  kable(
    dataFrame,
    format = "html",
    col.names = columns
  ) %>%
    # referredFrom: https://www.rdocumentation.org/packages/kableExtra/versions/1.4.0/topics/kable_styling
    kable_styling(
      # striped is used to get color contrast between rows.
      bootstrap_options = c("striped"),
      full_width = FALSE,
      position = "center"
    ) %>%
    # changing the header row.
    row_spec(
      0,
      bold = TRUE,
      color = "white",
      background = "#557A95"
    ) %>%
    # referredFrom: https://rdrr.io/cran/kableExtra/man/scroll_box.html
    scroll_box(
     extra_css = "border: none; padding: 0; max-height: 400px; overflow-y: auto;"
  )
}

# Function to get the top five products with the most sales/revenue across all years
# @param dbCon: database connection
getTopFiveProducts <- function(dbCon) {
  # Query to get the top five products with the most sales/revenue across all years
  query <- "SELECT pd.productName,ROUND(SUM(pf.totalRevenue),2) AS totalRevenue
FROM product_facts pf JOIN  product_dimension pd ON pf.productID = pd.productID
GROUP BY pd.productName
ORDER BY totalRevenue DESC LIMIT 5"
 return (dbGetQuery(dbCon, query))
}

displayReport(getTopFiveProducts(dbCon),c("Product Name", "Total Revenue"))
```

The total revenue per quarter per product for the most recent year is shown in the table below.

```{r totalRevenuePerQuarter, echo=FALSE,warning=FALSE, eval=TRUE}
# Function to get the total revenue per quarter per product for the most recent year
# @param dbCon: database connection
getTotalRevenuePerQuarter <- function(dbCon,recentYear) {
  # Query to get the total revenue per quarter per product for the most recent year
  query <- sprintf("SELECT
    pd.productName,
    pf.quarter,
    ROUND(SUM(pf.totalRevenue), 2) AS quaterlyRevenue
FROM
    product_facts pf JOIN product_dimension pd ON pf.productID = pd.productID
WHERE
    pf.year = %s
GROUP BY
    pd.productName,
    pf.quarter
ORDER BY
    productName,
    quarter", recentYear)
  return (dbGetQuery(dbCon, query))
}

displayReport(getTotalRevenuePerQuarter(dbCon,maxYear),c("Product Name", "Quarter", "Quarterly Revenue"))
```

The total revenue per product per country for the most recent two years is shown in the table below.

```{r totalRevenuePerProductPerCountry, echo=FALSE,warning=FALSE, eval=TRUE}
# Function to get the total revenue per product per country for the most recent two years
# @param dbCon: database connection
# @param recentYears: most recent two years
getTotalRevenuePerProductPerCountry <- function(dbCon, recentYears) {
  # Query to get the total revenue per product per country for the most recent two years
  query <- sprintf("SELECT pd.productName,
       cd.country,
       pf.year,
       ROUND(SUM(pf.totalRevenue), 2) AS totalRevenue
FROM product_facts pf
         JOIN product_dimension pd ON pf.productID = pd.productID
         JOIN customer_dimension cd ON pf.customerID = cd.customerID
WHERE pf.year IN (%s, %s)
GROUP BY pd.productName,
         cd.country,
         pf.year
ORDER BY pd.productName,
         cd.country,
         pf.year", recentYears[1], recentYears[2])
  return (dbGetQuery(dbCon, query))
}

displayReport(getTotalRevenuePerProductPerCountry(dbCon,c(yearInformation$year[1],yearInformation$year[2])),c("Product Name", "Country", "Year", "Total Revenue"))
```
The line graph below shows the sales per year for which we have data.

```{r salesPerYear, echo=FALSE,warning=FALSE, eval=TRUE}
# Function to visualize the trend of sales per year
# @param: dataFrame - data frame with data of sales per year
visualiseSalesByYear <- function(dataFrame) {
  # referredFrom: https://www.geeksforgeeks.org/r-line-graphs/
  plot(
    dataFrame$year,
    dataFrame$totalUnitsSold,
    # to plot lines and points
    type = "o",
    col = "darkblue",
    # to set filled circlers
    pch = 19,
    # to set line width
    lwd = 2,
    xlab = "Year",
    ylab = "Total Sales",
    main = "Total Sales by Year",
    ylim = c(0, max(dataFrame$total) * 1.3)
  )
  
  
  # to add grid lines
  # lty = 2 is used to set dashed lines
  grid(lty = 2, col = "gray80")
  
  # to add labels to the points
  text(
    # x location
    dataFrame$year,
    # y location
    dataFrame$totalUnitsSold + 500000,
    labels = dataFrame$totalUnitsSold,
    # size of text
    cex = 0.8,
    col = "darkblue",
  )

  
  # Adds a legend to the plot
  legend(
    "topleft",
    legend = "Total Sales",
    col = "darkblue",
    # used to denote a solid line
    lty = 1,
    # used to denote a filled circle
    pch = 19,
    bg = "white",
    # used to set box width
    box.lwd = 1
  )
}

getSalesByYear <- function(dbCon) {
  # Query to get the total sales per year
  query <- "SELECT
    year,
    SUM(totalUnitsSold) AS totalUnitsSold
FROM
    product_facts
GROUP BY
    year
ORDER BY
    year;"
  return (dbGetQuery(dbCon, query))
}

visualiseSalesByYear(getSalesByYear(dbCon))
```