---
title: "Assignment / Explore Query Planning and Indexing"
author: "Shubha, Adithya"
date: "2024-11-12"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r cleanEnvironment, echo=TRUE,warning=FALSE, eval= TRUE}
# Cleans the environment before beginning the execution
# referredFrom:https://northeastern.instructure.com/courses/192346/assignments/2351524
rm(list = ls())
```


```{r installAndLoadRequiredPackages, echo=TRUE,warning=FALSE, eval= TRUE}
# Install program required packages
# Installs RSQLite, DBI  packages
# referredFrom: http://artificium.us/lessons/06.r/l-6-104-r4progs/l-6-104.html#Install_Packages_on_Demand
installRequiredPackages <- function() {
  packages <- c("RSQLite", "DBI", "testthat")
  # Install packages that are not installed
  installed_packages <- packages %in% rownames(installed.packages())
  if (any(installed_packages == FALSE)) {
    install.packages(packages[!installed_packages])
  }
}

# Loads the required packages to the environment
# Loads RSQLite, DBI packages
loadRequiredPackages <- function() {
  suppressMessages({
    library(RSQLite)
    library(DBI)
    library(testthat)
  })
}
installRequiredPackages()
loadRequiredPackages()
```

```{r connectToDatabase, echo=TRUE,warning=FALSE, eval= TRUE}
# Connects to the database
# referredFrom: http://artificium.us/lessons/06.r/l-6-301-sqlite-from-r/l-6-301.html#Connect_to_Database
# @param dbName: Name of the database to connect to
connectToDatabase <- function(dbName) {
  return (dbConnect(RSQLite::SQLite(), dbname = dbName))
}

# Ensures the database connection is established
# by running a test query and asserting on the result.
# @param dbCon: Database connection object
ensureDatabaseConnection <- function(dbCon) {
  result <- dbGetQuery(dbCon, "SELECT * FROM film LIMIT 1")
  test_that("Result size is 1", {
    expect_equal(nrow(result), 1)
  })
  
}

# Function to clear databse cache to checking working of indexes
# @param dbCon: Database connection object
clearDatabaseCache <- function(dbCon) {
  # Free up cache memory in SQLite
  # referredFrom: https://stackoverflow.com/questions/4123196/sqlite-abnormal-memory-usage
  invisible(dbExecute(dbCon, "PRAGMA shrink_memory;"))
  
  # Garbage collection
  # referredFrom: https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/gc
  invisible(gc())
}

# Global database connection object
dbCon <- connectToDatabase("sakila.db")
ensureDatabaseConnection(dbCon)
clearDatabaseCache(dbCon)
```

## Question 1
```{r filmsPerCategory, eval=TRUE, echo=TRUE, warning=FALSE}

# Removes the given user defined index
# @param indexName: Name of the index to remove
queryToRemoveUserDefinedIndexes <- function(indexName) {
  return (sprintf("DROP INDEX IF EXISTS %s", indexName))
}

# Gets all user defined indexes
# @param dbCon: Database connection object
getAllUserDefinedIndexes <- function(dbCon) {
  return (dbGetQuery(dbCon, "SELECT name FROM sqlite_master WHERE type='index'"))
}

# Removes all user defined indexes
# @param dbCon: Database connection object
removeUserDefinedIndexes <- function(dbCon) {
  indexes <- getAllUserDefinedIndexes(dbCon)
  for (index in indexes$name) {
    # Ignoring auto generated indexes
    if (!grepl("sqlite_autoindex", index)) {
      dbExecute(dbCon, queryToRemoveUserDefinedIndexes(index))
    }
  }
}

# Query to get the film count per category
# Query assumes category_name and category_id have 1v1 mapping. 
queryToGetFilmCountPerCategory <- function() {
  return (
    "SELECT C.name as category_name, count(FC.film_id) as film_count
FROM film_category as FC
         JOIN category as C ON FC.category_id = C.category_id
GROUP BY C.name
"
  )
}

# Gets the film count per category
# @param dbCon: Database connection object
getFilmCountPerCategory <- function(dbCon) {
  result <- dbGetQuery(dbCon, queryToGetFilmCountPerCategory())
  
  # Formatting the film count to have a readable format
  result$film_count <- format(result$film_count,
                              scientific = FALSE,
                              big.mark = ",")
  return (result)
}

# Displays information on Zorro Ark film
# @param result: Result of the query
displayInformationOnFilm <- function(result) {
  print(result,row.names = FALSE)
}

# Displays the film count per category
# @param dbCon: Database connection object
displayFilmCountPerCategory <- function(dbCon) {
  result <- getFilmCountPerCategory(dbCon)
  displayInformationOnFilm(result)
}

removeUserDefinedIndexes(dbCon)
displayFilmCountPerCategory(dbCon)
```

## Question 2
```{r queryPlanForFilmsPerCategory, eval=TRUE, echo=TRUE, warning=FALSE}
# Gives the query plan for the given query
# @param dbCon: Database connection object
# @param query: Query for which the plan is required
getQueryPlan <- function(dbCon,query){
  return(dbGetQuery(dbCon, sprintf("EXPLAIN QUERY PLAN %s", query)))
}

# Displays the query plan for the given query
# Displays only the detail, other information can be added if required
# @param dbCon: Database connection object
# @param query: Query for which the plan is required
displayQueryPlan <- function(dbCon,query){
  result <- getQueryPlan(dbCon,query)
  print(result$detail,row.names = FALSE)
}

displayQueryPlan(dbCon,queryToGetFilmCountPerCategory())
```

## Question 3
```{r informationOnZorroArk, eval=TRUE, echo=TRUE, warning=FALSE}
# Query to get title,length, rental_rate, release_year of Zorro Ark film
queryToGetInformationOnZorroArk <- function() {
  return (
    "SELECT  title,length,rental_rate,release_year FROM film
where title = 'ZORRO ARK'")
}

getInformationOnZorroArk <- function(dbCon) {
  return (dbGetQuery(dbCon, queryToGetInformationOnZorroArk()))
}

clearDatabaseCache(dbCon)
queryStartTime <- Sys.time()
informationOnZorroArk <- getInformationOnZorroArk(dbCon)
queryEndTime <- Sys.time()
# in milliseconds
timeToFetchInformationOnZorroArk <- round(((queryEndTime - queryStartTime)*1000),3)
displayInformationOnFilm(informationOnZorroArk)
```

## Question 4
```{r queryPlanForZorroArk, eval=TRUE, echo=TRUE, warning=FALSE}
# display query plan on information of Zorro Ark
displayQueryPlan(dbCon,queryToGetInformationOnZorroArk())
```

## Question 5
```{r indexOnTitle, eval=TRUE, echo=TRUE, warning=FALSE}
# Query to create index on title column
queryToCreateIndexOnTitle <- function() {
  return ("CREATE INDEX IF NOT EXISTS TitleIndex
    ON film (title);")
}
createIndexOnTitle <- function(dbCon) {
  invisible(dbExecute(dbCon, queryToCreateIndexOnTitle()))
}

createIndexOnTitle(dbCon)
```

## Question 6
```{r queryPlanForZorroArkAfterIndex, eval=TRUE, echo=TRUE, warning=FALSE}
clearDatabaseCache(dbCon)
startTime <- Sys.time()
informationOnZorroArk <- getInformationOnZorroArk(dbCon)
endTime <- Sys.time()
# in milliseconds
timeToFetchInformationOnZorroArkAfterIndex <- round(((endTime - startTime)*1000),3)
displayInformationOnFilm(informationOnZorroArk)
displayQueryPlan(dbCon,queryToGetInformationOnZorroArk())
```

## Question 7
The query plan for question 4 and question 6 are different.
The difference is that query plan in question 6 uses the index to fetch the data while in 
question 4 the enitre film table is scanned to fetch the data.
We know question 6 is using index because the query plan shows the keyword 'USING INDEX' with the
index name 'TitleIndex' we created earlier.


## Question 8
The time taken to fetch information on Zorro Ark before creating the index is `r timeToFetchInformationOnZorroArk` ms and after creating the index is `r timeToFetchInformationOnZorroArkAfterIndex` ms
This shows that there is difference between both these times. The difference is
`r abs(timeToFetchInformationOnZorroArk - timeToFetchInformationOnZorroArkAfterIndex)` ms.
Ideally it should be the case that the time to fetch the information on Zorro Ark is reduced after creating the index on the title column since the index helps in fetching the data faster.

But if the data set is small then the time taken to fetch the data is almost the same before and after creating the index(as the difference is really low in milliseconds). 
This is because the time taken to scan the entire table is almost the same as the time taken to scan the index and fetch the data. Here index might introduce an overhead where it has to scan the index and then fetch the data from the table. This overhead might be more than the time taken to scan the entire table and fetch the data and is not worth if it is a single row lookup in a small table.  This is the reason why the time taken to fetch the data before and after creating the index is almost the same in-case of smaller dataset.


## Question 9
```{r informationWithFilmTitleHavingGold, echo=TRUE,warning=FALSE, eval= TRUE}
# Query to get information on films having title with gold
queryToGetInformationWithFilmTitleHavingGold <- function() {
  return (
    "SELECT 
    title,
    length,
    rental_rate,
    release_year
FROM film
WHERE LOWER(title) LIKE '%gold%';")
}

# Displays information on films having title with gold
# @param dbCon: Database connection object
displayInformationWithFilmTitleHavingGold <- function(dbCon) {
  result <- dbGetQuery(dbCon, queryToGetInformationWithFilmTitleHavingGold())
  displayInformationOnFilm(result)
}
displayInformationWithFilmTitleHavingGold(dbCon)
```

## Question 10
```{r queryPlanForFilmTitleHavingGold, echo=TRUE,warning=FALSE, eval= TRUE}
# Display query plan on information of films having title with gold
displayQueryPlan(dbCon,queryToGetInformationWithFilmTitleHavingGold())
```
The query plan shows that it is not using index. This happens because if an approximate search using LIKE is performed, then an index is not useful and the table is scanned row by row. The query plan above shows that the table is scanned and the index is ignored.

```{r disconnectFromDatabase, echo=TRUE,warning=FALSE, eval= TRUE}
# Disconnects from the database
# referredFrom: http://artificium.us/lessons/06.r/l-6-301-sqlite-from-r/l-6-301.html#Disconnect_from_Database
dbDisconnect(dbCon)
```